---
title: "Day 1A Practice"
format: html
---

![](../img/programmer_2780E3.svg){.hero}

## Question 1.

Read what the user wanted to do and look at the command they sent to R. Fix the errors in the commands and calculate the answers.

a)  I want to add 20 and 0.2 and divide that sum by 4.1. But I get an error when I try:<br />`[20 + .2] \ 4.1`
b)  I want to multiply 6 by the sum of 2 and 1,300. But I get an error when I try:<br />`6 x (2 + 1,300)`

<p><details>
<summary>Click here for the answer key</summary>
<blockquote>

**Answer (a)**

Change the brackets `[]` to parentheses `()` because parentheses `()` are the order of operations operator in R. Also, change the backslash `\` to a forward slash `/` because forward slash `/` is the division operator in R.

```{r}
#| comment: "#>"
#| collapse: true

(20 + .2) / 4.1
```

**Answer (b)**

Change the `x` to an asterisk `*` because asterisk `*` is the multiplication operator in R. Also, remove the comma `,` from `1,300` because R doesn't use commas to delimit digits in large numbers.

```{r}
#| comment: "#>"
#| collapse: true

6 * (2 + 1300)
```

</blockquote></details></p>

## Question 2.

Which of the following commands do you think will create an error in R? Why?

a) `score@T1 <- 3.2`
b) `score_at_T1 <- 3.2`
c) `score at T1 <- 3.2`
d) `1_score <- 3.2`
e) `ScoreAtTime1 <- 3.2`

Make your best guesses just by looking at the commands and then check your guesses by running the commands in R.

<p><details>
<summary>Click here for the answer key</summary>

> The following commands will create an error in R: **a** (`@` is not allowed in an object name), **c** (spaces are not allowed in an object name), and **d** (the first character in an object name cannot be a number).

</details></p>

## Resources

- [R4DS Chapter 8](https://r4ds.had.co.nz/workflow-projects.html): Read more about projects

- [R4DS Chapter 4](https://r4ds.had.co.nz/workflow-basics.html): Read about basic programming

- [R4DS Chapter 27](https://r4ds.had.co.nz/r-markdown.html): Read more about RMarkdown

## Fun Stuff

### Literal Genie

What it feels like to talk to R sometimes...

<iframe width="560" height="315" src="https://www.youtube-nocookie.com/embed/58l7ZqoU0a0" title="YouTube video player" frameborder="0" allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture" allowfullscreen></iframe>
