---
title: "Day 2A Practice"
format:
  html:
    df-print: paged
---

![](../img/maintenance_2780E3.svg){.hero}

```{r}
#| echo: false
#| message: false

library(tidyverse)
```

## Question 3.

Load the `economics` tibble (included in \{tidyverse\}). Then create a pipeline to accomplish the following data wrangling steps:

a)  First, in the `economics` tibble, rename the `uempmed` variable (the median duration of enemployment, in weeks) to `duration` and the `unemploy` variable (the number of unemployed, in thousands) to `number`.
b)  Then add a new variable called `rate` that is calculated by dividing `number` (from part a) by `pop` (the total population, in thousands).
c)  Then drop (i.e., unselect) the `pce`, `pop`, and `psavert` variables from the tibble.
d)  Then relocate the `rate` variable to be between the `date` and `duration` variables.

<p><details>
<summary>Click here for the answer key</summary>
<blockquote>

**Answer (a)**

```{r}
economics |> 
  rename(duration = uempmed, number = unemploy)
```

**Answer (b)**

```{r}
economics |> 
  rename(duration = uempmed, number = unemploy) |> 
  mutate(rate = number / pop)
```

**Answer (c)**

```{r}
economics |> 
  rename(duration = uempmed, number = unemploy) |> 
  mutate(rate = number / pop) |> 
  select(-c(pce, pop, psavert))
```

**Answer (d)**

```{r}
economics |>
  rename(duration = uempmed, number = unemploy) |> 
  mutate(rate = number / pop) |> 
  select(-c(pce, pop, psavert)) |> 
  relocate(rate, .after = date)
```

</blockquote>
</details></p>

## Question 3.

Download the [cereal](https://pittmethods.github.io/r4ss/data/cereal.csv) data file from the [workshop website](../Details/datasets.qmd). Copy it to your Project folder and read it into R as a tibble. Then create a pipeline to produce a short list of cereals for me to try. I am only interested in cold cereals with a rating greater than 70. Please arrange the list so that the cereals with the fewest calories are displayed at the top. Finally, write the result of this pipeline to a CSV file called "jeffs_list.csv" (I expect it to be even more popular than Craig's list).

<p><details>
<summary>Click here for the answer key</summary>
<blockquote>

```{r}
#| eval: false
#| message: false

cereal <- read_csv("cereal.csv") 

jeffs_list <- 
  cereal |> 
  filter(type == "cold" & rating > 70) |> 
  arrange(calories)

jeffs_list
```

```{r}
#| echo: false
#| message: false

cereal <- read_csv("../data/cereal.csv") 

jeffs_list <- 
  cereal |> 
  filter(type == "cold" & rating > 70) |> 
  arrange(calories)

jeffs_list
```

```{r}
#| eval: false

write_csv(jeffs_list, "jeffs_list.csv")
```

</blockquote>
</details></p>

## Question 1

a) Transform the `drv` variable in the `mpg` tibble into a factor where "4" is labeled Four Wheel Drive, "r" is labelled Rear Wheel Drive, and "f" is labeled Front Wheel Drive. Save this updated tibble as `mpg2`.

b) Transform the `manufacturer` and `model` variables in the `mpg2` tibble so that the first letter of each word is capitalized. Save this updated tibble as `mpg3`.

c) Create a new variable in the `mpg3` tibble called `name` which merges/unites the `manufacturer`, `model`, and `drv` variables into a single string (separated by spaces).

<p><details>
<summary>Click here for the answer key</summary>
<blockquote>

**Answer (a)**

```{r}
#| comment: "#>"
#| collapse: true

mpg2 <- 
  mpg |> 
  mutate(
    drv = factor(
      drv, 
      levels = c("4", "r", "f"), 
      labels = c("Four Wheel Drive", "Rear Wheel Drive", "Front Wheel Drive")
    )
  )
mpg2
```

**Answer (b)**

```{r}
#| comment: "#>"
#| collapse: true
#| warning: false

mpg3 <- 
  mpg2 |> 
  mutate(
    manufacturer = str_to_title(manufacturer),
    model = str_to_title(model)
  )
mpg3
```

**Answer (c)**

```{r}
#| comment: "#>"
#| collapse: true
#| warning: false

mpg3 |> 
  unite(col = "name", manufacturer, model, drv, sep = " ")
```

</blockquote></details></p>

## Question 2

a) Load the `txhousing` tibble (included in \{tidyverse\}). This contains the number of homes listed (`listings`) and the typical home sales price (`median`) in multiple Texas cities and months. Create a summary tibble that contains the average number of listings and the average typical home sales price across all cities and months. *Note:* For these descriptive purposes, you can ignore missing values.

b) Recreate your answer to Question 2a using `across()`.

c) Modify your answer to Question 2b to generate the same summary statistics but *per city* this time (still averaging across all months).

<p><details>
<summary>Click here for the answer key</summary>
<blockquote>

**Answer (a)**

```{r}
#| comment: "#>"
#| collapse: true

txhousing |> 
  summarize(
    listings = mean(listings, na.rm = TRUE), 
    median = mean(median, na.rm = TRUE)
  )

```

**Answer (b)**

```{r}
#| comment: "#>"
#| collapse: true
#| warning: false

txhousing |> 
  summarize(
    across(
      .cols = c(listings, median),
      .fns = mean, 
      na.rm = TRUE
    )
  )
```

```{r}
# Or, more compactly:

txhousing |> summarize(across(c(listings, median), mean, na.rm = TRUE))
```

**Answer (c)**

```{r}
#| comment: "#>"
#| collapse: true
#| warning: false

txhousing |> 
  group_by(city) |> 
  summarize(
    across(
      .cols = c(listings, median),
      .fns = mean, 
      na.rm = TRUE
    )
  )
```

</blockquote></details></p>

## Question 3

a) Use `if_else()` to add a new variable to the `txhousing` tibble called `crisis` which is equal to "pre" if `date` is less than 2007.5 and "post" otherwise.

b) Recreate your answer to Question 3a using `case_when()`.

c) *Bonus:* Add a new variable to the `txhousing` tibble called `season` which is equal to: "spring" when `month` is 3, 4, or 5; "summer" when `month` is 6, 7, or 8; "autumn" when `month` is 9, 10, or 11; and "winter" when `month` is 1, 2, or 12.

<p><details>
<summary>Click here for the answer key</summary>
<blockquote>

**Answer (a)**

```{r}
#| comment: "#>"
#| collapse: true

txhousing |> 
  mutate(
    crisis = if_else(
      condition = date < 2007.5,
      true = "pre",
      false = "post"
    )
  )
```

**Answer (b)**

```{r}
#| comment: "#>"
#| collapse: true
#| warning: false

txhousing |> 
  mutate(
    crisis = case_when(
      date < 2007.5 ~ "pre",
      date >= 2007.5 ~ "post"
    )
  )
```

**Answer (c)**

```{r}
#| comment: "#>"
#| collapse: true
#| warning: false

txhousing |> 
  mutate(
    season = case_when(
      month %in% c(3, 4, 5) ~ "spring",
      month %in% c(6, 7, 8) ~ "summer",
      month %in% c(9, 10, 11) ~ "autumn",
      month %in% c(12, 1, 2) ~ "winter"
    )
  )
```

</blockquote></details></p>

## Resources

-   [R4DS Chapter 5](https://r4ds.had.co.nz/transform.html) (§5.1–§5.5): Read more about the basic wrangling verbs 

- [R4DS Chapter 15](https://r4ds.had.co.nz/factors.html): Read more about working with factors in R

## Fun Stuff

### Stunt Rope

These are some advanced wrangling verbs...

<iframe width="560" height="315" src="https://www.youtube-nocookie.com/embed/flcFl8z_hF4?start=99" title="YouTube video player" frameborder="0" allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture" allowfullscreen></iframe>
