---
format: 
  revealjs:
    css: ../styles.css
    slide-number: true
    show-slide-number: all
    preview-links: auto
    self-contained: true
    progress: true
    history: true
    hash-type: number
    theme: default
    code-block-background: true
    highlight-style: zenburn
    code-link: false
    code-copy: true
    pagetitle: "R4SS Day 2A"
    author-meta: "Jeffrey Girard"
    date-meta: "2022-07-26"
---

::: {.my-title}
# [Introduction to R]{.blue} <br />for Social Scientists

::: {.my-grey}
[Workshop Day 2A | 2022-07-26]{}<br />
[Jeffrey M. Girard | Pitt Methods]{}
:::

![](../img/proud_coder_357EDD.svg){.absolute bottom=0 right=0 width=400}
:::

<!-- Program II -->

# Program II

## Missing Values {.smaller}

::: {.columns .pv4}
::: {.column width="60%"}
-   Sometimes your data will have [missing values]{.b .blue}
    -   Perhaps these were [never collected]{.b .green}
    -   Perhaps the values were [lost/corrupted]{.b .green}
    -   Perhaps the participant [didn't respond]{.b .green}

::: {.fragment .mt1}
-   We need to tell R which values are missing
    -   To do so, we set those values to `NA`
    -   Functions from **tidyverse** make this easy
:::

::: {.fragment .mt1}
-   Missingness is often ["contagious"]{.b .green} in R<br />
    e.g., a vector with `NA` has an unknown mean
:::
:::

::: {.column .tc .pv5 width="40%"}
{{< li rwotyanb trigger=loop delay=3000 speed=0.5 colors=secondary:#2a76dd class=rc >}}
:::
:::

::: footer
\[2A\] Program II
:::

## Missing Values Live Coding

```{r}
#| echo: true
#| eval: false
#| error: true
#| code-line-numbers: false

# SETUP: We will need tidyverse for the read and mutate functions

library(tidyverse)

# ==============================================================================

# PITFALL: Number codes for missingness will mess up calculations in R

heights <- c(149, 158, -999) # here we use -999 to represent a missing value

range(heights)

mean(heights)

log(heights) # our missing value is no longer -999

# ==============================================================================

# USECASE: Use NA for missingness instead

heights2 <- c(149, 158, NA)
heights2

log(heights2) # the NA stayed an NA (due to contagiousness)

# ==============================================================================

# LESSON: Use na.rm = TRUE to do a summary function ignoring the NAs

mean(heights2) # the mean is an NA (due to contagiousness)

mean(heights2, na.rm = TRUE)

range(heights2, na.rm = TRUE)

# ==============================================================================

# USECASE: Dealing with missing values in tibbles

cereal <- read_csv("cereal.csv")

cereal$rating

range(cereal$rating)

# ==============================================================================

# LESSON: Use na_if() to convert specific values to NA while mutating

cereal2 <- mutate(cereal, rating = na_if(rating, -999))

cereal2$rating

range(cereal2$rating, na.rm = TRUE)

# ==============================================================================

# LESSON: Use read_csv(na) to convert specific values to NA while reading

cereal3 <- read_csv("cereal.csv", na = "-999")

cereal3$rating

range(cereal3$rating, na.rm = TRUE)
```

::: footer
\[2A\] Program II
:::

## Factors {.smaller}

::: {.columns .pv4}
::: {.column width="60%"}
-   [Factors]{.b .blue} are used to represent categorical data
    -   Factors have multiple possible [levels]{.b .blue}
    -   Levels are [discrete]{.b .green} and mutually-exclusive

::: {.fragment .mt1}
-   Sometimes categories are [unordered]{.b .green} (nominal)
    -   Action or Comedy or Drama
    -   Asia or Europe or North America
:::

::: {.fragment .mt1}
-   Sometimes categories are [ordered]{.b .green} (ordinal)
    -   Mild < Medium < Hot
    -   XS < S < M < L < XL
:::
:::

::: {.column .tc .pv5 width="40%"}
{{< li cdbgwqyw trigger=loop delay=3000 colors=secondary:#2a76dd class=rc >}}
:::
:::

::: footer
\[2A\] Program II
:::

## Factors Live Coding

```{r}
#| echo: true
#| eval: false
#| error: true
#| code-line-numbers: false

# USECASE: Ask 10 kids to order 1: nuggets, 2: pizza, or 3: salad

food <- c(2, 2, 1, 2, 1, 2, 1, 1, 2, 2)
food

# ==============================================================================

# LESSON: We can turn this vector into a factor with factor()

food2 <- factor(food, levels = c(1, 2, 3))
food2

food3 <- factor(food, levels = c(1, 2, 3),
                labels = c("nuggets", "pizza", "salad"))
food3

# ==============================================================================

# USECASE: We can also quickly and easily count each level with table()

table(food3)

# ==============================================================================

# PITFALL: Don't confuse levels and labels

food4 <- factor(food, labels = c(1, 2, 3),
                levels = c("nuggets", "pizza", "salad"))
food4 # full of <NA> because it can't find these levels

# ==============================================================================

# USECASE: You can also just enter strings directly (as self-labels)

genre <- c("pop", "metal", "pop", "rock", "rap", "rap", "pop", "rock")
genre

genre2 <- factor(genre) # observed levels will be assigned alphabetically
genre2

table(genre2)

# ==============================================================================

# LESSON: If ordinal, enter levels low-to-high and add ordered = TRUE

salsa <- c("hot", "mild", "medium", "mild", "medium", "medium")

salsa2 <- factor(salsa, 
                 levels = c("mild", "medium", "hot"), 
                 ordered = TRUE)
salsa2 

# NOTE: We may want to visualize or model ordinal factors differently

# ==============================================================================

# USECASE: Working with factors in a tibble

cereal <- read_csv("cereal.csv")
cereal

cereal2 <- mutate(cereal, mfr = factor(mfr), type = factor(type))
cereal2

table(cereal2$mfr)

table(cereal2$type)
```

::: footer
\[2A\] Program II
:::

## String Manipulation {.smaller}

::: {.columns .pv4}
::: {.column width="60%"}
-   It can be helpful to [transform strings]{.b .blue}
    -   Sometimes your [data]{.b .green} is text-based
    -   Text also appears in [tables]{.b .green} and [figures]{.b .green}...
    -   So we may want to customize it quickly

::: {.fragment .mt1}
-   Tidyverse contains tools for this
    -   The functions all begin with `str_`
    -   We will only discuss the basics
    -   Learn more at [stringr.tidyverse.org](https://stringr.tidyverse.org/)
:::

:::

::: {.column .tc .pv5 width="40%"}
{{< li puvaffet trigger=loop delay=3000 colors=secondary:#2a76dd class=rc >}}
:::
:::

::: footer
\[2A\] Program II
:::

## String Manipulation Live Coding

```{r}
#| echo: true
#| eval: false
#| error: true
#| code-line-numbers: false

# SETUP: We will need tidyverse for almost all of these functions

library(tidyverse)

# ==============================================================================

# USECASE: Re-capitalize strings

x <- c("R4SS: Introduction to R for Social Scientists")

str_to_lower(x)

str_to_upper(x)

str_to_sentence(x)

str_to_title(x)

# ==============================================================================

# USECASE: Extract subsets of strings

x <- c("Apple", "Banana", "Pear")

str_sub(x, start = 1, end = 3)

str_sub(x, start = -3, end = -1)

str_sub(x, start = 2, end = -2)

str_sub(x, start = 1, end = 5) # can go beyond the end

# ==============================================================================

# USECASE: Remove whitespace from strings

x <- "  Sometimes strings have   too   much white space "
x

str_trim(x) # remove white space at the start and end

str_squish(x) # trim and then collapse inner white space

# ==============================================================================

# USECASE: Remove and replace patterns in strings

x <- "Scientists very often utilize very fancy words, 
      even when they could utilize simpler ones."

str_remove(x, pattern = "very ") # removes first pattern match only

str_remove_all(x, pattern = "very ") # removes all pattern matches

str_replace(x, pattern = "utilize", replacement = "use")

str_replace_all(x, pattern = "utilize", replacement = "use")

# NOTE: More complex patterns can be found using regular expressions (regex)

# ==============================================================================

# USECASE: Create a string manipulation pipeline

x_clean <- 
  x |> 
  str_remove_all("very ") |> 
  str_replace_all("utilize", "use") |> 
  print()
```

::: footer
\[2A\] Program II
:::

## If Else {.smaller}

::: {.columns .pv4}
::: {.column width="60%"}
-   A [locked door]{.b .green} behaves [conditionally]{.b .blue}
    -   If you have the key, then open up...
    -   Otherwise, stay closed...

::: {.fragment .mt1}
-   Sometimes we want [code]{.b .blue} to behave conditionally
    -   *Filter* retains observations conditionally<br />
        (e.g., if it meets a condition, it gets to stay)
    -   Let's learn to [transform]{.b .green} variables conditionally
    -   We can use `if_else()` for simple examples
:::

:::

::: {.column .tc .pv5 width="40%"}
{{< li efoxgafi trigger=loop delay=3000 colors=secondary:#2a76dd class=rc >}}
:::
:::

::: footer
\[2A\] Program II
:::

## If Else Live Coding

```{r}
#| echo: true
#| eval: false
#| error: true
#| code-line-numbers: false

# SETUP: We will need tidyverse for almost all of these functions

library(tidyverse)

# ==============================================================================

# USECASE: Determining whether someone can vote in the US

age <- 12

age_group <- 
  if_else(
    condition = age >= 18,
    true = "adult",
    false = "child"
  ) |> 
  print()

# ==============================================================================

# TIP: Because argument names are optional, we can shorten this (if we want)

age_group <- if_else(age >= 18, "adult", "child") |> print()

# ==============================================================================

# LESSON: This function is particularly useful applied to vectors

ages <- c(13, 18, 14, 19, 22, 16)

age_groups <- if_else(ages >= 18, "adult", "child") |> print()

# ==============================================================================

# USECASE: We can therefore use it during data wrangling

cereal <- read_csv("cereal.csv", na = "-999")

cereal2 <- mutate(cereal, popular = if_else(rating > 50, "yes", "no"))
cereal2

cereal3 <- 
  cereal |>
  mutate(
    diabetes = if_else(
      condition = sugars == 0, 
      true = "sugar-free", 
      false = "contains sugar"
    )
  ) |> 
  print()
```

::: footer
\[2A\] Program II
:::

## Case When {.smaller}

::: {.columns .pv4}
::: {.column width="60%"}
-   An [elevator]{.b .green} also behaves [conditionally]{.b .blue}
    -   If you press a button, then it goes to that floor
    -   There are usually more than just two buttons
    -   In this analogy (but not in real life), the elevator only responds to the [first]{.b .green} button pressed

::: {.fragment .mt1}
-   Sometimes we want [code]{.b .blue} to behave this way
    -   `case_when()` expands upon `if_else()`
    -   It can have [multiple conditions]{.b .blue} (floor buttons)
    -   The first condition met "wins" (picks the floor)
:::
:::

::: {.column .tc .pv5 width="40%"}
{{< li nwyovgla trigger=loop delay=3000 colors=secondary:#2a76dd class=rc >}}
:::
:::

::: footer
\[2A\] Program II
:::

## Case When Live Coding

```{r}
#| echo: true
#| eval: false
#| error: true
#| code-line-numbers: false

# SETUP: We will need tidyverse for almost all of these functions

library(tidyverse)

# ==============================================================================

# USECASE: Determine what types of movies your kids can watch

ages <- c(11, 13, 18)

movies_allowed <- 
  case_when(
    ages >= 17 ~ "R",
    ages >= 13 ~ "PG-13",
    ages < 13 ~ "PG"
  ) |> 
  print()

# ==============================================================================

# PITFALL: Don't put the least restrictive condition first

age <- 18

movies_allowed2 <- 
  case_when(
    age < 13 ~ "PG",
    age >= 13 ~ "PG-13",
    age >= 17 ~ "R"
  ) |> 
  print() # age >= 13, so PG-13 wins before checking if age >= 17

# ==============================================================================

# USECASE: Use case_when to re-code variables during data wrangling

starwars

sw <- 
  starwars |> 
  mutate(
    species3 = case_when(
      species == "Human" ~ "Human",
      species == "Droid" ~ "Droid",
      species != "Human" & species != "Droid" ~ "Alien"
    )
  ) |> 
  select(name, species3) |> 
  print()

# ==============================================================================

# TIP: The next version of case_when() will add the .default argument
# This is where the elevator will drop you off if you hit no buttons

sw <- 
  starwars |> 
  mutate(
    species3 = case_when(
      species == "Human" ~ "Human",
      species == "Droid" ~ "Droid",
      .default = "Alien"
    )
  ) |> 
  select(name, species3) |> 
  print()

# NOTE: The above code won't work now, but it should in a few weeks/months
# For now, you can use TRUE ~ "Alien" and it works but is harder to explain
```

::: footer
\[2A\] Program II
:::

<!-- Wrangle III -->

# Wrangle III

## Summarize {.smaller}

::: {.columns .pv4}
::: {.column width="60%"}
-   Although we store data about many observations...
-   ...we often want to [summarize across]{.b .blue} observations
    -   This is like [folding]{.b .green} the tibble down to [one row]{.b .green}

::: {.fragment .mt1}
-   We've seen functions that summarize vectors
    -   `length()`, `sum()`, `min()`, `max()`
    -   `mean()`, `median()`, `sd()`, `var()`
:::

::: {.fragment .mt1}
-   `summarize()` lets us use them on tibbles
    -   It works very similarly to `mutate()`
    -   It always creates a tibble as output
:::
:::

::: {.column .tc .pv5 width="40%"}
{{< li uhmqrsgl trigger=loop delay=3000 colors=secondary:#2a76dd class=rc >}}
:::
:::

::: footer
\[2A\] Wrangle III
:::

## Summarize Live Coding

```{r}
#| echo: true
#| eval: false
#| error: true
#| code-line-numbers: false

# SETUP: We will need tidyverse and an example dataset

library(tidyverse)

sales <- 
  tibble(
    customer = c(1, 2, 3, 1, 3),
    store = c("A", "A", "A", "B", "B"),
    items = c(25, 20, 16, 10, 5),
    spent = c(685, 590, 392, 185, 123)
  ) |> 
  print()

# ==============================================================================

# USECASE: Summarize the typical sales

my_summary <- 
  sales |> 
  summarize(
    avg_items = mean(items),
    avg_spent = mean(spent)
  ) |> 
  print()

# ==============================================================================

# PITFALL: Don't use summary() instead of summarize()

my_summary <- 
  sales |> 
  summary(
    avg_items = mean(items),
    avg_spent = mean(spent)
  ) |> 
  print()

# ==============================================================================

# USECASE: Use more than one summary function

my_summary <- 
  sales |> 
  summarize(
    total_items = sum(items),
    total_spent = sum(spent),
    avg_items = mean(items),
    avg_spent = mean(spent)
  ) |> 
  print()

# ==============================================================================

# USECASE: Use counting functions

my_counts <- 
  sales |> 
  summarize(
    n_sales = n(),
    n_customers = n_distinct(customer),
    n_stores = n_distinct(store)
  ) |> 
  print()
```

::: footer
\[2A\] Wrangle III
:::

## Group Summarize {.smaller}

::: {.columns .pv4}
::: {.column width="60%"}
- We can also summarize a tibble [by group]{.b .blue}
    -   This is like [folding]{.b .green} the tibble multiple times
    -   Specifically, we fold down to [one row per group]{.b .green}

::: {.fragment .mt1}
-   The syntax for summarize is identical
    -   The only difference is to the tibble
    -   We first pass it through `group_by()`
    -   Pipelines make this very easy
:::
:::

::: {.column .tc .pv5 width="40%"}
{{< li pmwtbpap trigger=loop delay=3000 colors=secondary:#2a76dd class=rc >}}
:::
:::

::: footer
\[2A\] Wrangle III
:::

## Group Summarize Live Coding

```{r}
#| echo: true
#| eval: false
#| error: true
#| code-line-numbers: false

# SETUP: We will need tidyverse and an example dataset

library(tidyverse)

sales <- 
  tibble(
    customer = c(1, 2, 3, 1, 3),
    store = c("A", "A", "A", "B", "B"),
    items = c(25, 20, 16, 10, 5),
    spent = c(685, 590, 392, 185, 123)
  ) |> 
  print()

# ==============================================================================

# LESSON: We pass a tibble through group_by to group it

sales

sales |> group_by(store) # note the display says "grouped"

# ==============================================================================

# USECASE: We can then summarize and get stats per group

sales |> 
  group_by(store) |> 
  summarize(
    customers = n_distinct(customer),
    items_sold = sum(items),
    total_sales = sum(spent),
    avg_items = mean(items),
    avg_spent = mean(spent)
  )

# ==============================================================================

# SETUP: Let's get a larger, more realistic dataset

# Extra pane > Packages tab > Install > nycflights13

library("nycflights13")

flights

# ==============================================================================

# USECASE: Find the carrier with the lowest average delays

flights |> 
  group_by(carrier) |> 
  summarize(m_delay = mean(dep_delay, na.rm = TRUE)) |> 
  arrange(m_delay)

# ==============================================================================

# LESSON: We can also group by multiple variables

# USECASE: Let's find the day of the year with the most flights

flights |> 
  group_by(month, day) |> 
  summarize(n_flights = n()) |> 
  arrange(desc(n_flights))

# ==============================================================================

# PITFALL: Note how this differs from grouping by just day

flights |> 
  group_by(day)|> 
  summarize(n_flights = n()) |> 
  arrange(desc(n_flights))
```

::: footer
\[2A\] Wrangle III
:::

## Across {.smaller}

::: {.columns .pv4}
::: {.column width="60%"}
-   We can use `across()` to [repeat an operation]{.b .blue} across multiple variables in a tibble
    -   This makes our code [shorter]{.b .green}
    -   It is [faster]{.b .green} to read and write
    -   It is also [less error-prone]{.b .green}

::: {.fragment .mt1}
-   So we can repeat a function in order to...
    -   ...`mutate()` multiple variables
    -   ...`summarize()` multiple variables
:::
:::

::: {.column .tc .pv5 width="40%"}
{{< li iectdupg trigger=loop delay=3000 colors=secondary:#2a76dd class=rc >}}
:::
:::

## Across Live Coding

```{r}
#| echo: true
#| eval: false
#| error: true
#| code-line-numbers: false

# SETUP: We will need tidyverse and an example dataset

library(tidyverse)

starwars

# ==============================================================================

# USECASE: Applying the same mutation to multiple variables is a pain

sw <- 
  starwars |> 
  mutate(
    hair_color = factor(hair_color),
    skin_color = factor(skin_color),
    eye_color = factor(eye_color)
  ) |> 
  print() # before

sw <- 
  starwars |> 
  mutate(
    across(
      .cols = c(hair_color, skin_color, eye_color), 
      .fns = factor
    )
  ) |> 
  print() #after

# ==============================================================================

# PITFALL: Don't forget to wrap the .cols part in c()

sw <- 
  starwars |> 
  mutate(
    across(
      .cols = mass, birth_year, 
      .fns = round,
      digits = 1
    )
  ) |> 
  print() # error

# ==============================================================================

# LESSON: To pass arguments to the inner function, add them inside across()

sw <- 
  starwars |> 
  mutate(
    across(
      .cols = c(mass, birth_year), 
      .fns = round,
      digits = 1
    )
  ) |> 
  print()

# ==============================================================================

# USECASE: You can also apply the same summary functions across variables

sw <- 
  starwars |> 
  summarize(
    height = mean(height, na.rm = TRUE),
    mass = mean(mass, na.rm = TRUE),
    birth_year = mean(birth_year, na.rm = TRUE)
  ) |> 
  print()

sw <- 
  starwars |> 
  summarize(
    across(
      .cols = c(height, mass, birth_year), 
      .fns = mean, 
      na.rm = TRUE
    )
  ) |> 
  print()
```

::: footer
\[2A\] Wrangle III
:::

## Separate and Unite {.smaller}

::: {.columns .pv4}
::: {.column width="60%"}
-   Tidy data needs one value per cell

::: {.fragment .mt1}
-   So we may need to [separate]{.b .blue} cells
    -   *e.g., What was the model of my first car?*
    -   `"Nissan Altima 2003"` ⬎
    -   `"Nissan"` `"Altima"` `"2003"`
:::

::: {.fragment .mt1}
-   But some tasks require us to [unite]{.b .blue} cells
    -   *e.g., What address should I mail to?*
    -   `123` `"Main Street"` ⬎
    -   `"123 Main Street"`
:::
:::

::: {.column .tc .pv5 width="40%"}
{{< li dsfanfaz trigger=loop delay=3000 colors=secondary:#2a76dd class=rc >}}
:::
:::

::: footer
\[2A\] Wrangle III
:::

## Separate Live Coding

```{r}
#| echo: true
#| eval: false
#| error: true
#| code-line-numbers: false

# SETUP: We will need tidyverse and an example dataset

library(tidyverse)

# Create some example data
dat <- 
  tibble(
    id = c("A_001_01", "A_002_01", "B_001_01", "B_002_01", "C_001_01", "C_002_01"),
    duration = c("01:16", "01:21", "01:49", "00:34", "00:32", "00:54")
  ) |> 
  print()

# ==============================================================================

# USECASE: Separate a column into multiple columns

dat2 <- 
  dat |> 
  separate(
    col = duration, 
    into = c("min", "sec"), 
    sep = ":"
  ) |> 
  print()

# ==============================================================================

# USECASE: This also works with more than two "into" columns

dat2 <- 
  dat |>  
  separate(
    col = id, 
    into = c("group", "subject", "time"), 
    sep = "_"
  ) |> 
  print()

# ==============================================================================

# TIP: To automatically convert strings into numbers, use convert

dat2 <- 
  dat |> 
  separate(
    col = id, 
    into = c("group", "subject", "time"), 
    sep = "_", 
    convert = TRUE
  ) |> 
  print()

# ==============================================================================

# PITFALL: Don't forget to close the "into" vector's parentheses

dat2 <- 
  dat|> 
  separate(col = duration, into = c("min", "sec", sep = ":")) #error
```

::: footer
\[2A\] Wrangle III
:::

## Unite Live Coding

```{r}
#| echo: true
#| eval: false
#| error: true
#| code-line-numbers: false

# SETUP: We will need tidyverse and an example dataset

library(tidyverse)

dat2 <- 
  dat|> 
  separate(col = id, into = c("group", "subject", "time"), sep = "x") |> 
  separate(col = duration, into = c("min", "sec"), sep = ":") |> 
  print()

# ==============================================================================

# USECASE: Unite multiple columns into one string

dat3 <- 
  dat2 |> 
  unite(col = "newid", group, subject, time, sep = "-") |> 
  unite(col = "duration", min, sec, sep = ":") |> 
  print()

# ==============================================================================

# LESSON: Retain the columns being united with remove = FALSE

dat3 <- 
  dat2 |> 
  unite(col = "newid", group:time, sep = "", remove = FALSE) |> 
  print()
```

::: footer
\[2A\] Wrangle III
:::

## Pivot Longer and Wider {.smaller}

::: {.columns .pv4}
::: {.column width="60%"}
-   Both long and wide formats can be tidy
    -   Long formats are better for MLM
    -   Wide formats are better for SEM

::: {.fragment .mt1}
-   It can be useful to quickly [reshape a tibble]{.b .blue}
    -   `pivot_longer()`: wide → long
    -   `pivot_wider()`: long → wide
:::
:::

::: {.column .tc .pv5 width="40%"}
{{< li xacejjis trigger=loop delay=3000 speed=0.5 colors=secondary:#2a76dd class=rc >}}
:::
:::

## Pivot Longer Live Coding

```{r}
#| echo: true
#| eval: false
#| error: true
#| code-line-numbers: false

# SETUP: We will need tidyverse and an example dataset (from workshop website)

library(tidyverse)

gradebook <- 
  read_csv("gradebook.csv") |> 
  print()

# ==============================================================================

# USECASE: We can pivot to long format by creating name and value variables

gradebook2 <- 
  gradebook |> 
  pivot_longer(
    cols = c(test1, test2, test3, test4, test5), 
    names_to = "test", 
    values_to = "grade"
  ) |> 
  print()

# ==============================================================================

# TIP: Use selection helpers to select columns quickly

gradebook2 <- 
  gradebook |> 
  pivot_longer(
    cols = test1:test5, 
    names_to = "test", 
    values_to = "grade"
  ) |> 
  print()

# ==============================================================================

# LESSON: Automatically remove the name prefix

gradebook2 <- 
  gradebook|> 
  pivot_longer(
    cols = starts_with("test"), 
    names_to = "test", 
    values_to = "grade",
    names_prefix = "test"
  ) |> 
  print()
```

::: footer
\[2A\] Wrangle III
:::

## Pivot Wider Live Coding

```{r}
#| echo: true
#| eval: false
#| error: true
#| code-line-numbers: false

# SETUP: We will need tidyverse and an example dataset (from workshop website)

library(tidyverse)

diary <- read_csv("diary.csv") |> print()

# ==============================================================================

# USECASE: Reshape this long format to a wider format

diary_scale <- 
  diary |> 
  pivot_wider(
    names_from = "scale",
    values_from = "score"
  ) |> 
  print()

diary_day <- 
  diary |>  
  pivot_wider(
    names_from = "day",
    values_from = "score"
  ) |> 
  print()

# NOTE: There are thus multiple possible wide formats (for different uses)

# ==============================================================================

# LESSON: We can add a prefix to each name to avoid numeric names

diary_data

diary_day <- 
  diary |> 
  pivot_wider(
    names_from = "day",
    values_from = "score",
    names_prefix = "day_"
  ) |> 
  print()

# ==============================================================================

# LESSON: We can also pivot on multiple columns at once

diary_double <- 
  diary |> 
  pivot_wider(
    names_from = c("scale", "day"),
    values_from = "score"
  ) |> 
  print()
```

::: footer
\[2A\] Wrangle III
:::

<!-- Practice III -->

# [Practice III](https://pittmethods.github.io/r4ss/Day_2/Day2A_Practice.html){preview-link="false"}
