---
title: "Datasets"
format:
  html:
    df-print: paged
---

```{r}
#| echo: false
#| message: false

library(tidyverse)
```


## Datasets to download

### affairs

Real data on 601 participants' extramarital activities.

<a href="https://pittmethods.github.io/r4ss/data/affairs.csv"><button type="button" class="btn btn-primary">Download CSV</button></a>

```{r}
#| echo: false
#| message: false

read_csv("../data/affairs.csv")
```

### cereal

Real data on 77 cereal products from the early 1990s. \[[Source](https://www.kaggle.com/datasets/crawford/80-cereals)\]

<a href="https://pittmethods.github.io/r4ss/data/cereal.csv"><button type="button" class="btn btn-primary">Download CSV</button></a>

```{r}
#| echo: false
#| message: false

read_csv("../data/cereal.csv")
```

### diary

Real data from 112 participants in a "daily diary" study on stress and affect. \[[Source](https://doi.org/10.1521/pedi.2015.29.4.503)\]

<a href="https://pittmethods.github.io/r4ss/data/diary.csv"><button type="button" class="btn btn-primary">Download CSV</button></a>

```{r}
#| echo: false
#| message: false

read_csv("../data/diary.csv")
```

### ds_salaries

Real data about 607 data scientist's jobs and salaries during 2020-2022. \[[Source](https://www.kaggle.com/datasets/saurabhshahane/data-science-jobs-salaries)\]

<a href="https://pittmethods.github.io/r4ss/data/ds_salaries.csv"><button type="button" class="btn btn-primary">Download CSV</button></a>

```{r}
#| echo: false
#| message: false

read_csv("../data/ds_salaries.csv")
```

### elicitation

Real data about 139 participants' emotions during four emotion elicitation tasks. \[[Source](https://doi.org/10.1016/j.imavis.2014.06.002)\]

<a href="https://pittmethods.github.io/r4ss/data/elicitation.csv"><button type="button" class="btn btn-primary">Download CSV</button></a>

```{r}
#| echo: false
#| message: false

read_csv("../data/elicitation.csv")
```

### exercise

Simulated data about 900 participants' weight loss under different exercise programs.

<a href="https://pittmethods.github.io/r4ss/data/exercise.csv"><button type="button" class="btn btn-primary">Download CSV</button></a>

```{r}
#| echo: false
#| message: false

read_csv("../data/exercise.csv")
```

### gradebook

Simulated data on the scores of 10 students on 5 tests.

<a href="https://pittmethods.github.io/r4ss/data/gradebook.csv"><button type="button" class="btn btn-primary">Download CSV</button></a>

```{r}
#| echo: false
#| message: false

read_csv("../data/gradebook.csv")
```

### interpersonal

Real data about the interpersonal problems of 34 psychotherapy clients. \[[Source](https://doi.org/10.1109/FG52635.2021.9667021)\]

<a href="https://pittmethods.github.io/r4ss/data/interpersonal.csv"><button type="button" class="btn btn-primary">Download CSV</button></a>

```{r}
#| echo: false
#| message: false

read_csv("../data/interpersonal.csv")
```

### political

Real data about the political activities of 59 college students.

<a href="https://pittmethods.github.io/r4ss/data/political.csv"><button type="button" class="btn btn-primary">Download CSV</button></a>

```{r}
#| echo: false
#| message: false

read_csv("../data/political.csv")
```

### salaries

Real data about 397 professors during 2008 to 2009. \[[Source](https://cran.r-project.org/package=carData)\]

<a href="https://pittmethods.github.io/r4ss/data/salaries.csv"><button type="button" class="btn btn-primary">Download CSV</button></a>

```{r}
#| echo: false
#| message: false

read_csv("../data/salaries.csv")
```

### yearspubs

<a href="https://pittmethods.github.io/r4ss/data/yearspubs.csv"><button type="button" class="btn btn-primary">Download CSV</button></a>

```{r}
#| echo: false
#| message: false

read_csv("../data/yearspubs.csv")
```

## Datasets in Packages

### economics 

Real economic time series data about the US from 1967 to 2015. \[[Source](https://fred.stlouisfed.org/)\]

Included in the \{ggplot2\} package and loaded by \{tidyverse\}.

```{r}
#| echo: false

economics
```

### flights

Real data about all 336,766 flights in and out of New York City in 2013. \[[Source](https://www.transtats.bts.gov/DL_SelectFields.asp?Table_ID=236)\]

Included in the \{nycflights13\} package.

```{r}
#| echo: false

nycflights13::flights
```

### mpg

Real data about 234 car models from 1999 and 2008. \[[Source](https://fueleconomy.gov/)\]

Included in the \{ggplot2\} package and loaded by \{tidyverse\}.

```{r}
#| echo: false

mpg
```

### penguins

Real data about 344 penguins observed in Palmer Archipelago. \[[Source](https://doi.org/10.1371/journal.pone.0090081)\]

Included in the \{palmerpenguins\} package.

```{r}
#| echo: false

palmerpenguins::penguins
```

### starwars

Real data about 87 characters in the Star Wars franchise. \[[Source](https://swapi.dev)\]

Included in the \{dplyr\} package and loaded by \{tidyverse\}.

```{r}
#| echo: false

starwars
```

### txhousing

Real data about the Texas housing market from 2000 to 2015. \[[Source](https://www.recenter.tamu.edu/)\]

Included in the \{ggplot2\} package and loaded by \{tidyverse\}.

```{r}
#| echo: false

txhousing
```

