---
title: "Day 3B Practice"
format:
  html:
    df-print: paged
---

![](../img/growth_curve_2780E3.svg){.hero}

```{r}
#| echo: false
#| message: false

library(tidyverse)
```

## Question 1.

a) Download the [political](https://pittmethods.github.io/r4ss/data/political.csv) dataset from the [workshop website](https://pittmethods.github.io/r4ss/Details/datasets.html), copy it to your project folder, and read it into R. Mutate this tibble such that the `sex` variable is a factor (0=Male, 1=Female). 

b) Create a histogram to view the distributions of the `informed` and `tv_news` variables and a bar plot to view the counts of the `sex` variable. What do these distributions look like?

c) Build a regression model to predict each student's self-reported level of political informedness (`informed`) from the degree to which they watch television news (`tv_news`). What is the standardized slope for the television news predictor? *Bonus:* What is the model's adjusted R-squared?

d) Expand your regression model from Question 1(b) to become a multiple regression model that also includes students' sex (`sex`) as a predictor. What are the two predictor variables' partial effects?

e) Statistically test whether the effect of television watching depends on students' sex and generate a plot that shows the model's predictions (to help you interpret any interaction effect).

<p><details>
<summary>Click here for the answer key</summary>
<blockquote>

**Answer (a)**

```{r}
#| comment: "#>"
#| eval: false
#| collapse: true

library(tidyverse)
political <- 
  read_csv("political.csv") |> 
  mutate(sex = factor(sex, levels = c(0, 1), labels = c("Male", "Female"))) |> 
  print()
```

```{r}
#| comment: "#>"
#| echo: false
#| collapse: true
#| message: false

library(tidyverse)
political <- 
  read_csv("../data/political.csv") |> 
  mutate(sex = factor(sex, levels = c(0, 1), labels = c("Male", "Female"))) |> 
  print()
```

**Answer (b)**

```{r}
#| comment: "#>"
#| collapse: true
#| warning: false
#| message: false

ggplot(political, aes(x = informed)) + geom_histogram()
ggplot(political, aes(x = tv_news)) + geom_histogram()
ggplot(political, aes(x = sex)) + geom_bar()
```

The `informed` variable is discrete at 1, 2, 3, 4, or 5 with a triangular pattern where 3 is most common. The `tv_news` variable is discreet at 0, 1, 2, 3, or 4 with a pattern where 0 and 3 are most common. The `sex` variable is relatively balanced between male and female.

**Answer (c)**

```{r}
#| comment: "#>"
#| collapse: true
#| warning: false

library(parameters)
library(performance)
fit <- lm(informed ~ tv_news, data = political)
model_parameters(fit, standardized = "refit")
model_performance(fit)
```

The standardized slope is 0.21 for `tv_news` and it is significantly different from zero $(p=.004)$. This simple model explains 12% of the variance in `informed` $(R^2_{adj}=0.123)$.

**Answer (d)**

```{r}
#| comment: "#>"
#| collapse: true
#| warning: false

fit2 <- lm(informed ~ tv_news + sex, data = political)
model_parameters(fit2)
```

The partial effect of television news was $0.20$ and the partial effect of sex was $-0.44$ (females reported feeling/being less informed than males).

**Answer (e)**

```{r}
#| comment: "#>"
#| collapse: true
#| warning: false

library(ggeffects)
fit3 <- lm(informed ~ tv_news * sex, data = political)
model_parameters(fit3)
ggeffect(fit3, terms = c("tv_news", "sex")) |> plot()
```

The effect of television news does not seem to depend on sex, as the interaction effect was not significant $(p=.887)$ and the prediction lines were very close to parallel.

</blockquote></details></p>

## Question 2

Did you bring your own data? If so, load it into R and explore it using the tools we learned this week. Wrangle it a bit and maybe do some basic statistics on it, estimating a few regression models. If not, then look through the datasets on the [workshop website](https://pittmethods.github.io/r4ss/Details/datasets.html) and find one that is interesting to you; do the same process I described above to practice your skills in a more self-directed manner.
