---
title: "Day 3A Practice"
format:
  html:
    df-print: paged
---

![](../img/statistics_2780E3.svg){.hero}

```{r}
#| echo: false
#| message: false

library(tidyverse)
```

## Question 1.

a) Download the [yearspubs](https://pittmethods.github.io/r4ss/data/yearspubs.csv) dataset from the [workshop website](https://pittmethods.github.io/r4ss/Details/datasets.html), copy it to your project folder, and read it into R.

b) Estimate the correlation between academics' number of publications (`n_pubs`) and number of citations (`n_cites`) in this sample.

c) Estimate the average number of citations for male and female academics (with 95\% confidence intervals).

d) Estimate the effect size of this difference and test whether it is statistically significant. You can decide whether to use a parametric or nonparametric approach.

<p><details>
<summary>Click here for the answer key</summary>
<blockquote>

**Answer (a)**

```{r}
#| comment: "#>"
#| eval: false
#| collapse: true

library(tidyverse)
yearspubs <- read_csv("yearspubs.csv")
yearspubs
```

```{r}
#| comment: "#>"
#| echo: false
#| collapse: true
#| message: false

yearspubs <- read_csv("../data/yearspubs.csv")
yearspubs
```

**Answer (b)**

```{r}
#| comment: "#>"
#| collapse: true
#| warning: false

library(correlation)
correlation(yearspubs)
```

The correlation between number of publications and citations is 0.34.

**Answer (c)**

```{r}
#| comment: "#>"
#| collapse: true
#| warning: false

library(statsExpressions)
centrality_description(yearspubs, x = sex, y = n_cites)
```

The average was estimates as 37.3 \[33.4, 43.5\] for females and 42.4 \[35.0, 48.7\] for males.

**Answer (d)**

```{r}
#| comment: "#>"
#| collapse: true
#| warning: false

two_sample_test(yearspubs, x = sex, y = n_cites, type = "parametric")
```

The effect size was $g=-0.3$ but the difference was not significant $(p=.232)$.

```{r}
#| message: false

# Optionally:
library(ggstatsplot)
ggbetweenstats(yearspubs, x = sex, y = n_cites)
```


</blockquote></details></p>

## Question 2

Using the `mpg` tibble from the \{tidyverse\} package, statistically test whether cars with the three different drive trains (`drv`) get significantly different city fuel efficiency (`cty`) on average. Calculate the effect size for this difference.

<p><details>
<summary>Click here for the answer key</summary>
<blockquote>

```{r}
#| comment: "#>"
#| collapse: true
#| warning: false

oneway_anova(mpg, x = drv, y = cty)
```

The effect size was $\omega^2=0.68$ and the difference was significant $(p<.001)$.

```{r}
#| comment: "#>"
#| collapse: true
#| warning: false

ggbetweenstats(mpg, x = drv, y = cty)
```

</blockquote></details></p>

## Question 3

Did you bring your own data? If so, load it into R and explore it using the tools we learned this week. Wrangle it a bit and maybe do some basic statistics on it, estimating correlations and comparing group means. If not, then look through the datasets on the [workshop website](https://pittmethods.github.io/r4ss/Details/datasets.html) and find one that is interesting to you; do the same process I described above to practice your skills in a more self-directed manner.

## Resources

-   [Web](https://easystats.github.io/correlation/articles/types.html): Read more about the methods offered by the \{correlation\} package

-   [Web](https://indrajeetpatil.github.io/statsExpressions/articles/statsExpressions.html#expressions-for-plots): Read more about all the information in a \{ggstatsplot\} graphic

-   [Learning Statistics with R](https://learningstatisticswithr.com/): A free textbook on the ideas/theory underlying the statistical models we just discussed
