---
format: 
  revealjs:
    css: ../styles.css
    slide-number: true
    show-slide-number: all
    preview-links: auto
    #self-contained: true
    progress: true
    history: true
    hash-type: number
    theme: default
    code-block-background: true
    highlight-style: zenburn
    code-link: false
    code-copy: true
    pagetitle: "R4SS Day 3A"
    author-meta: "Jeffrey Girard"
    date-meta: "2022-07-27"
---

::: {.my-title}
# [Introduction to R]{.blue} <br />for Social Scientists

::: {.my-grey}
[Workshop Day 3A | 2022-07-27]{}<br />
[Jeffrey M. Girard | Pitt Methods]{}
:::

![](../img/proud_coder_357EDD.svg){.absolute bottom=0 right=0 width=400}
:::

<!-- Model I -->

# Model I

## Data Verification {.smaller}

::: {.columns .pv4}
::: {.column width="60%"}
-   It's always a good idea to start with [verification]{.b .blue}

::: {.fragment .mt1}
-   Check that your variables are the [correct type]{.b .green}
    -   Configure your [factors]{.b .green}' levels and labels
    -   Establish ordinal factors' ordering
    -   Explicitly set your [missing values]{.b .green} to NA
:::

::: {.fragment .mt1}
-   Check variables' [extrema]{.b .green} and [distributions]{.b .green}
    -   Check for erroneous and outlying values
    -   Check the shape of continuous distributions
    -   Check the overlap of categorical levels
:::
:::

::: {.column .tc .pv5 width="40%"}
{{< li zqwicaxf trigger=loop delay=3000 colors=secondary:#2a76dd class=rc >}}
:::
:::

::: footer
\[3A\] Model I
:::

## Data Verification Live Coding

```{r}
#| echo: true
#| eval: false
#| error: true
#| code-line-numbers: false

# SETUP:

library(tidyverse)
library(datawizard)

salaries <- read_csv("salaries.csv")

# ==============================================================================

# LESSON: First, set all your categorical variables to factors

salaries

salaries <- 
  salaries |> 
  mutate(across(c(rank, discipline, sex), factor)) |> 
  print()

# ==============================================================================

# LESSON: Then check the summary statistics for problems

summary(salaries)

describe_distribution(salaries)

data_tabulate(salaries, exclude = is.numeric)

# ==============================================================================

# LESSON: Then visualize the variable's distributions

ggplot(salaries, aes(x = rank)) + geom_bar()

ggplot(salaries, aes(x = discipline)) + geom_bar()

ggplot(salaries, aes(x = sex)) + geom_bar()

ggplot(salaries, aes(x = yrs.since.phd)) + geom_histogram()

ggplot(salaries, aes(x = yrs.service)) + geom_histogram()

ggplot(salaries, aes(x = salary)) + geom_histogram()

# ==============================================================================

# LESSON: Finally, check for empty level-combinations

salaries |> 
  group_by(rank, discipline, sex) |> 
  summarize(n = n()) |> 
  arrange(n)
```

::: footer
\[3A\] Model I
:::

## Correlations {.smaller}

::: {.columns .pv4}
::: {.column width="60%"}
-   [Correlations]{.b .blue} $(r)$ quantify the strength of the [linear relationship]{.b .green} between two variables from $-1$ to $+1$
    -   $r\rightarrow-1$: as $x$ increases, $y$ decreases
    -   $r\rightarrow0$: as $x$ increases, $y$ doesn't change
    -   $r\rightarrow+1$: as $x$ increases, $y$ also increases

::: {.fragment .mt1}
-   Correlations may be the focus of statistical inference or just useful descriptive summaries
:::

::: {.fragment .mt1}
-   We can easily estimate many [different types]{.b .green} of correlation coefficients in R
:::
:::

::: {.column .tc .pv5 width="40%"}
{{< li nbdmfygb trigger=loop delay=3000 colors=secondary:#2a76dd class=rc >}}
:::
:::

::: footer
\[3A\] Model I
:::

## Correlations Live Coding

```{r}
#| echo: true
#| eval: false
#| error: true
#| code-line-numbers: false

# SETUP: Load the used packages (if needed) and read in the example dataset

library(tidyverse)
library(correlation)

salaries <- read_csv("salaries.csv")

# ==============================================================================

# LESSON: The standard correlation test in R

cor.test(salaries$salary, salaries$yrs.since.phd)

# ==============================================================================

# TIP: Fancier correlations with the {correlation} package (from {easystats})

correlation(salaries)

# ==============================================================================

# LESSON: Creating a graphical model plot (requires installing some packages)

correlation(salaries) |> plot()

# ==============================================================================

# LESSON: Creating and plotting the correlation matrix

correlation(salaries) |> summary()

correlation(salaries) |> summary() |> plot()

correlation(salaries) |> summary(redundant = TRUE) |> plot()

# ==============================================================================

# p-value adjustments

correlation(salaries, p_adjust = "none") # very liberal

correlation(salaries, p_adjust = "bonferroni") # very conservative

correlation(salaries, p_adjust = "holm") # recommended

# ==============================================================================

# LESSON: Other correlation methods

correlation(salaries, method = "kendall") # rank correlation

correlation(salaries, method = "blomqvist") # similar to kendall but better

correlation(salaries, method = "distance") # linear and nonlinear relationships
```

::: footer
\[3A\] Model I
:::

## Comparing Two Groups {.smaller}

::: {.columns .pv4}
::: {.column width="60%"}
-   A fundamental task in science is comparing the [centrality]{.b .green} of two groups' distributions
    -   It is common to compare means or medians

::: {.fragment .mt1}
-   [Independent]{.b .blue} groups are [separate]{.b .green}
    -   Comparisons are [between]{.b .green} subjects
    -   *Did students in New York or students in California spend more time on social media?*
:::

::: {.fragment .mt1}
-   [Dependent]{.b .blue} groups are [paired]{.b .green}
    -   Comparisons are [within]{.b .green} subjects
    -   *Did the students in my class spend more time on social media during the winter or the summer?*
:::
:::

::: {.column .tc .pv5 width="40%"}
{{< li wfkxiwtw trigger=loop delay=3000 colors=secondary:#2a76dd class=rc >}}
:::
:::

::: footer
\[3A\] Model I
:::

## Independent Groups Live Coding

```{r}
#| echo: true
#| eval: false
#| error: true
#| code-line-numbers: false

# SETUP: Load the used packages (if needed) and read in the example dataset

library(tidyverse)
library(statsExpressions)
library(ggstatsplot)

salaries <- read_csv("salaries.csv")

# ==============================================================================

# USECASE: Let's compare the salaries across the two disciplines

two_sample_test(
  data = salaries, 
  x = discipline, 
  y = salary, 
  type = "parametric"
)

# ==============================================================================

# LESSON: We can also estimate each group's means by swapping the function

centrality_description(
  data = salaries,
  x = discipline, 
  y = salary, 
  type = "parametric"
)

# ==============================================================================

# LESSON: Nonparametric does not assume a normal distribution of y

two_sample_test(
  data = salaries, 
  x = discipline, 
  y = salary, 
  type = "nonparametric"
)

centrality_description(
  data = salaries, 
  x = discipline, 
  y = salary, 
  type = "nonparametric"
)

# ==============================================================================

# TIP: Calculate the test results and generate a plot using {ggstatsplot}

ggbetweenstats(
  data = salaries, 
  x = discipline, 
  y = salary, 
  type = "parametric"
)

ggbetweenstats(
  data = salaries, 
  x = discipline, 
  y = salary, 
  type = "nonparametric"
)
```

::: footer
\[3A\] Model I
:::

## Dependent Groups Live Coding

```{r}
#| echo: true
#| eval: false
#| error: true
#| code-line-numbers: false

# SETUP: Load the used packages (if needed) and read in the example dataset

library(tidyverse)
library(statsExpressions)
library(ggstatsplot)

interpersonal <- read_csv("interpersonal.csv")
interpersonal

# ==============================================================================

# LESSON: Be sure to set paired to TRUE and provide a subject.id

two_sample_test(
  data = interpersonal,
  x = time,
  y = problems,
  paired = TRUE,
  subject.id = patient,
  type = "parametric"
)

centrality_description(
  data = interpersonal,
  x = time,
  y = problems,
  type = "parametric"
)

# ==============================================================================

# LESSON: There is a nonparametric version of this too

two_sample_test(
  data = interpersonal,
  x = time,
  y = problems,
  paired = TRUE,
  subject.id = patient,
  type = "nonparametric"
)

centrality_description(
  data = interpersonal,
  x = time,
  y = problems,
  type = "nonparametric"
)

# ==============================================================================

# TIP: No need to add paired because we are switching to ggwithinstats()

ggwithinstats(
  data = interpersonal, 
  x = time, 
  y = problems,
  subject.id = patient,
  type = "parametric"
)

ggwithinstats(
  data = interpersonal, 
  x = time, 
  y = problems,
  subject.id = patient,
  type = "nonparametric"
)
```

## Comparing Many Groups {.smaller}

::: {.columns .pv4}
::: {.column width="60%"}
-   We may also compare [more than two]{.b .green} groups
    -   Independent vs. dependent still applies

::: {.fragment .mt1}
-   An [omnibus test]{.b .blue} asks *whether* groups are different
    -   If significant, one or more group is different
:::

::: {.fragment .mt1}
-   [Posthoc tests]{.b .blue} ask *how* groups are different
    -   Pairwise tests compare each pair of groups
    -   They are often "gated" behind omnibus tests
    -   With many groups, *p*-values are adjusted
:::
:::

::: {.column .tc .pv5 width="40%"}
{{< li dugfwxlj trigger=loop delay=3000 colors=secondary:#2a76dd class=rc >}}
:::
:::

::: footer
\[3A\] Model I
:::

## Many Independent Live Coding

```{r}
#| echo: true
#| eval: false
#| error: true
#| code-line-numbers: false

# SETUP: Load the used packages (if needed) and read in the example dataset

library(tidyverse)
library(statsExpressions)
library(ggstatsplot)

salaries <- read_csv("salaries.csv")

# ==============================================================================

# USECASE: Let's compare the salaries across the three ranks

oneway_anova(
  data = salaries,
  x = rank,
  y = salary,
  type = "parametric"
)

centrality_description(
  data = salaries,
  x = rank,
  y = salary,
  type = "parametric"
)

oneway_anova(
  data = salaries,
  x = rank,
  y = salary,
  type = "nonparametric"
)

# ==============================================================================

# USECASE: Let's compare the salaries between each pair of ranks

pairwise_comparisons(
  data = salaries,
  x = rank,
  y = salary,
  type = "parametric"
)

pairwise_comparisons(
  data = salaries,
  x = rank,
  y = salary,
  type = "nonparametric"
)

# ==============================================================================

# TIP: Do it all (overall and pairwise tests) with ggbetweenstats

ggbetweenstats(
  data = salaries, 
  x = rank, 
  y = salary,
  type = "parametric"
)

ggbetweenstats(
  data = salaries, 
  x = rank, 
  y = salary,
  type = "nonparametric"
)
```

::: footer
\[3A\] Model I
:::

## Many Dependent Live Coding

```{r}
#| echo: true
#| eval: false
#| error: true
#| code-line-numbers: false

# SETUP: Load the used packages (if needed) and read in the example dataset

library(tidyverse)
library(statsExpressions)
library(ggstatsplot)

elicitation <- read_csv("elicitation.csv")
elicitation

# ==============================================================================

# USECASE: Let's compare the amusement self-reports across the four tasks

oneway_anova(
  data = elicitation,
  x = Task,
  y = Amused,
  paired = TRUE,
  subject.id = Subject,
  type = "parametric"
)

centrality_description(
  data = elicitation,
  x = Task,
  y = Amused,
  type = "parametric"
)

oneway_anova(
  data = elicitation,
  x = Task,
  y = Amused,
  paired = TRUE,
  subject.id = Subject,
  type = "nonparametric"
)

# ==============================================================================

# USECASE: Let's compare the amusement self-reports between each pair of tasks

pairwise_comparisons(
  data = elicitation,
  x = Task,
  y = Amused,
  paired = TRUE,
  subject.id = Subject,
  type = "parametric"
)

pairwise_comparisons(
  data = elicitation,
  x = Task,
  y = Amused,
  paired = TRUE,
  subject.id = Subject,
  type = "nonparametric"
)

# ==============================================================================

# TIP: Do it all (overall and pairwise tests) with ggwithinstats

ggwithinstats(
  data = elicitation,
  x = Task,
  y = Amused,
  paired = TRUE,
  subject.id = Subject,
  type = "parametric"
)

ggwithinstats(
  data = elicitation,
  x = Task,
  y = Amused,
  paired = TRUE,
  subject.id = Subject,
  type = "nonparametric"
)
```

::: footer
\[3A\] Model I
:::

<!-- Model II -->

# Model II 

## Basic Regression

::: {.columns .pv4}
::: {.column width="60%"}
A

::: {.fragment .mt1}
B
:::

::: {.fragment .mt1}
C
:::
:::

::: {.column .tc .pv5 width="40%"}
{{< li ogfgksuz trigger=loop delay=3000 colors=secondary:#2a76dd class=rc >}}
:::
:::

::: footer
\[3A\] Model II
:::

## Basic Regression Live Coding

```{r}
#| echo: true
#| eval: false
#| error: true
#| code-line-numbers: false

# SETUP: Load the used packages (if needed) and read in the example dataset

library(tidyverse)
library(datawizard)
library(parameters)
library(performance)
library(ggeffects)

salaries <- read_csv("salaries.csv")
salaries

# ==============================================================================

fit <- lm(
  formula = salary ~ yrs.since.phd,
  data = salaries
)

fit

model_parameters(fit)

model_parameters(fit, standardize = "refit")

model_performance(fit)

plot(ggpredict(fit, terms = "yrs.since.phd"))

plot(ggpredict(fit, terms = "yrs.since.phd"), add.data = TRUE)
```

::: footer
\[3A\] Model II
:::

## Multiple Regression

::: {.columns .pv4}
::: {.column width="60%"}
A

::: {.fragment .mt1}
B
:::

::: {.fragment .mt1}
C
:::
:::

::: {.column .tc .pv5 width="40%"}
{{< li jmkpuued trigger=loop delay=3000 colors=secondary:#2a76dd class=rc >}}
:::
:::

::: footer
\[3A\] Model II
:::

## Multiple Regression Live Coding

```{r}
#| echo: true
#| eval: false
#| error: true
#| code-line-numbers: false


```

::: footer
\[3A\] Model II
:::

## Categorical Predictors

::: {.columns .pv4}
::: {.column width="60%"}
A

::: {.fragment .mt1}
B
:::

::: {.fragment .mt1}
C
:::
:::

::: {.column .tc .pv5 width="40%"}
{{< li rwotyanb trigger=loop delay=3000 colors=secondary:#2a76dd class=rc >}}
:::
:::

::: footer
\[3A\] Model II
:::

## Categorical Live Coding

```{r}
#| echo: true
#| eval: false
#| error: true
#| code-line-numbers: false


```

::: footer
\[3A\] Model II
:::

## Interaction Effects

::: {.columns .pv4}
::: {.column width="60%"}
A

::: {.fragment .mt1}
B
:::

::: {.fragment .mt1}
C
:::
:::

::: {.column .tc .pv5 width="40%"}
{{< li nmlpnruz trigger=loop delay=3000 colors=secondary:#2a76dd class=rc >}}
:::
:::

::: footer
\[3A\] Model II
:::

## Interactions Live Coding

```{r}
#| echo: true
#| eval: false
#| error: true
#| code-line-numbers: false


```

::: footer
\[3A\] Model II
:::

## Polynomial Regression

::: {.columns .pv4}
::: {.column width="60%"}
A

::: {.fragment .mt1}
B
:::

::: {.fragment .mt1}
C
:::
:::

::: {.column .tc .pv5 width="40%"}
{{< li lcellbfh trigger=loop delay=3000 colors=secondary:#2a76dd class=rc >}}
:::
:::

::: footer
\[3A\] Model II
:::

## Polynomial Live Coding

```{r}
#| echo: true
#| eval: false
#| error: true
#| code-line-numbers: false


```

::: footer
\[3A\] Model II
:::

## Regression Diagnostics

::: {.columns .pv4}
::: {.column width="60%"}
A

::: {.fragment .mt1}
B
:::

::: {.fragment .mt1}
C
:::
:::

::: {.column .tc .pv5 width="40%"}
{{< li oamdefle trigger=loop delay=3000 colors=secondary:#2a76dd class=rc >}}
:::
:::

::: footer
\[3A\] Model II
:::

## Diagnostics Live Coding

```{r}
#| echo: true
#| eval: false
#| error: true
#| code-line-numbers: false


```

::: footer
\[3A\] Model II
:::

<!-- Practice V -->

# [Practice V](https://pittmethods.github.io/r4ss/Day_3/Day3A_Practice.html){preview-link="false"}
